# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
Load activity.csv file with a header
```{r}
activity <- read.csv("activity.csv", header=TRUE, na.strings="NA")
```

Clean up the data by converting internal to factor, and date to a Date
```{r}
##Convert interval to factor
activity$interval <- factor(activity$interval)
##Convert date to Date
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
```


## What is mean total number of steps taken per day?
Create a data frame that shows the total amout of steps per day
```{r}
## Create steps vector
steps <- activity$steps
## Create a dates vector to group by
date <- activity$date
## Create data frame grouped by day
stepsByDay <- as.data.frame(rowsum(steps, date,na.rm=TRUE))
##Add Day as column and convert to Date
stepsByDay$date <- rownames(stepsByDay)
stepsByDay$date <- as.Date(stepsByDay$date, format="%Y-%m-%d")
colnames(stepsByDay) <- c("steps", "date")
```

Plot date in data frame using a histogram
```{r Frequency of Steps by Day, message=FALSE}
library("ggplot2")
ggplot(stepsByDay, aes(x=steps)) + geom_histogram() +
        ggtitle("Frequency of Number of Steps by Day")
```

Calculate mean and median steps per day
```{r}
##Produce a data frame with mean by day
meanSteps <- aggregate(steps ~ date, activity, mean, na.action=NULL)
meanSteps$date <- as.character(meanSteps$date)
colnames(meanSteps) <- c("date", "mean steps")
##Produce a data frame with median by day
medianSteps <- aggregate(steps ~ date, activity, median, na.action=NULL)
medianSteps$date <- as.character(medianSteps$date)
colnames(medianSteps) <- c("date", "median steps")
```

###Table of Mean(Left) and Median(Right) By Day
```{r, results='asis'}
library(knitr)
t1 = kable(meanSteps, format='html', output = FALSE)
t2 = kable(medianSteps, format='html', output = FALSE)
cat(c('<table><tr valign="top"><td>', t1, '</td><td>', t2, '</td><tr></table>'),
    sep = '')
```


## What is the average daily activity pattern?
###Plot  of the 5-minute interval and the average number of steps taken, 
###averaged across all days (y-axis)
```{r Average Number of Steps by Interval, fig.width=20, fig.height=10}
meanStepsByInterval <- aggregate(steps ~ interval, activity, mean, 
        na.action=na.omit)
##Set mean steps as numeric and clean up column names
colnames(meanStepsByInterval) <- c("interval", "Average_Steps")
meanStepsByInterval$interval <- as.numeric(meanStepsByInterval$interval)
ggplot(meanStepsByInterval, aes(interval, Average_Steps)) + geom_line() +
    ggtitle("Average Number of Steps by Interval")
```

Calculate the interval with the highest average (accross all days)
```{r}
maxInterval <- which.max(meanStepsByInterval$Average_Steps)
meanStepsByInterval[maxInterval,]
```

## Imputing missing values

The number of NAs is:
```{r}
length(which(is.na(activity$steps)))
```

Replace missing NA values with the mean of their interval:
```{r, warning=FALSE}
##Convert interval to numeric
activity$interval <- as.numeric(activity$interval)
##Merge average steps by Interval into activity data frame
activityFill <- merge(activity, meanStepsByInterval, by="interval")
##Replace NAs with mean of interval
activityFill$steps[is.na(activityFill$steps)] <- activityFill$Average_Steps
```

Calculate mean and median steps per day
```{r}
##Produce a data frame with mean by day
meanStepsFill <- aggregate(steps ~ date, activityFill, mean, na.action=NULL)
meanStepsFill$date <- as.character(meanStepsFill$date)
colnames(meanStepsFill) <- c("date", "mean steps")
##Produce a data frame with median by day
medianStepsFill <- aggregate(steps ~ date, activityFill, median, 
        na.action=NULL)
medianStepsFill$date <- as.character(medianStepsFill$date)
colnames(medianStepsFill) <- c("date", "median steps")
```

###Show original mean table (left) vs NA Filled (right)
```{r two-tables-mean, results='asis'}
t1 = kable(meanSteps, format='html', output = FALSE)
t2 = kable(meanStepsFill, format='html', output = FALSE)
cat(c('<table><tr valign="top"><td>', t1, '</td><td>', t2, '</td><tr></table>'),
    sep = '')
```

###Show original median table (left) vs NA Filled (right)
```{r two-tables-median, results='asis'}
t1 = kable(medianSteps, format='html', output = FALSE)
t2 = kable(medianStepsFill, format='html', output = FALSE)
cat(c('<table><tr valign="top"><td>', t1, '</td><td>', t2, '</td><tr></table>'),
    sep = '')
```
  
  
###Do these values differ from the estimates from the first part of the assignment?  
###What is the impact of imputing missing data on the estimates of the total daily
###number of steps?  
Yes, there were clearly entire missing days of values which affected both  
mean and median measures for some days.  This shows a clear impact of imputing
missing data.


## Are there differences in activity patterns between weekdays and weekends?
Add weekdays to imputed/filled data frame and add new factor Weekday/Weekend
```{r}
activityFill$weekday <- weekdays(activityFill$date)
##Add Weekday to daycat column
activityFill$daycat[activityFill$weekday == "Monday" | 
        activityFill$weekday == "Tuesday" | 
        activityFill$weekday == "Wednesday" | 
        activityFill$weekday == "Thursday" | 
        activityFill$weekday == "Friday"] <- "Weekday"
##Add Weekend to daycat column
activityFill$daycat[activityFill$weekday == "Saturday" | 
        activityFill$weekday == "Sunday"] <- "Weekend"
##Convert daycat to factor
activityFill$daycat <- factor(activityFill$daycat)
```

###Plot  of the 5-minute interval and the average number of steps taken, 
###averaged across all weekend days and all weekday days (y-axis)
```{r Weekday vs Weekday Interval Steps}
##Create weekday data frame with mean steps by interval
meanStepsByIntervalWeekday <- aggregate(steps ~ interval, subset(activityFill,
        daycat == "Weekday"), mean, na.action=na.omit)
##Set mean steps for Weekday as numeric and clean up column names
colnames(meanStepsByIntervalWeekday) <- c("interval", "Average_Steps")
meanStepsByIntervalWeekday$interval <- 
        as.numeric(meanStepsByIntervalWeekday$interval)

##Create weekend data frame with mean steps by interval
meanStepsByIntervalWeekend <- aggregate(steps ~ interval, subset(activityFill,
        daycat == "Weekend"), mean, na.action=na.omit)
##Set mean steps for Weekend as numeric and clean up column names
colnames(meanStepsByIntervalWeekend) <- c("interval", "Average_Steps")
meanStepsByIntervalWeekend$interval <- 
        as.numeric(meanStepsByIntervalWeekend$interval)

library(grid)
library(gridExtra)
p1 <- ggplot(meanStepsByIntervalWeekday, aes(interval, Average_Steps)) + 
        geom_line() + ggtitle("Average Number of Steps by Interval for Weekday Days")
p2 <- ggplot(meanStepsByIntervalWeekend, aes(interval, Average_Steps)) + 
        geom_line() + ggtitle("Average Number of Steps by Interval for Weekend Days")
grid.arrange(p1, p2, ncol = 1, main = "Weekday vs Weekday Interval Steps")
```